Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit

### 2024-04-12 optim.qmd
smoothness measure
Update of the smoothness measure part with equations

The jellyfish optimiser (JSO) mimics the natural movements of jellyfish,
which include passive and active motions driven by ocean currents and
their swimming patterns, respectively. In the context of optimization,
these movements are abstracted to explore the search space in a way that
balances exploration (searching new areas) and exploitation (focusing on
promising areas). The algorithm aims to find the optimal solution by
adapting the jellyfish's behavior to navigate towards the best solution
over iterations [@chou_novel_2021].

To understand what the jellyfish optimizer is doing in the context of
Projection Pursuit, we first start with a current projection (the
starting point). Then we evaluate this projection using an index
function, which tells us how good the current projection is. We then
move the projection in a direction determined by the 'best jelly' and
random factors, influenced by how far along we are in the optimization
process (the trial $i$ and `max.tries`). Occasionally, we might explore
completely new directions like a jellyfish might with ocean currents.
Then, we compare new potential projections to our current one. If
they're better, we adopt them; if not, we stick with our current
projection. This process continues and iteratively improves the
projection, until we reach the maximum number of trials.

::: {.callout-note icon="false"}

## Algorithm: Jellyfish Optimizer Pseudo Code

@@

>

@@

> **if** $c_t$ is greater than or equal to $0.5$ **then**
>    
> > Define trend based on the best jelly and current projections
>     
> > Update each projection towards the trend using a random factor and orthonormalisation
>       
> **else**
>     
> > **if** a random number is greater than $1 - c_t$ **then**
> >       
> > > Slightly adjust each projection with a small random factor (Type A passive)
> >         
> > **else**
> >       
> > > For each projection, compare with a random jelly and adjust towards or away from it (Type B active)
>         
> Update the orientation of each projection to maintain consistency
>     
> Evaluate the new projections using the index function

>

> **if** any new projection is worse than the current, revert to the `current_projection` for that case
> 
> > Determine the projection with the best index value as the new best_jelly

>

> **if** the try is the last one, print the final best projection and **exit**

**return** the set of projections with the updated best_jelly as the optimized_projection

:::

@@

<!--
An intuitive way to measure smoothness of a function would be to count
how many continuous derivatives exist. 
To help define smoothness in our
context, we can make use of the Sobolev spaces: functions $f$ in the
Sobolev space $W^{p,\infty}$ have all derivatives of order less than $p$
continuous. Smoothness would then be the highest $p$ such that the index
function belongs to $W^{p,\infty}$. We can relax this and consider
$W^{p,q}$ Sobolev spaces.

@@

-   $h_i$ is an neighbourhood area around $a$. We can insert different
    values of $h$ to approximate the above quantity and observe how this
    quantity changes.
-->

If we evaluate the index function at some random points (like the random initialization of the jellyfish optimizer), then we can interpret these random index values as a random field, indexed by a space parameter: the random projection angle. This analogy suggests to use this random training sample to fit a spatial model, a simple one being a (spatial) Gaussian process. There exist R packages to fit the hyperparameters of a Gaussian process, for example ExaGeoStatR (https://github.com/ecrc/exageostatR), which is an R wrapper for the ExaGeoStat library (https://arxiv.org/pdf/1708.02835.pdf).

How can we define a measure of smoothness from this? A Gaussian process is fully defined by its mean function and covariance function. The way the covariance function is defined is where smoothness comes into play: if an index is very smooth, then two close projection angles should produce close index values (strong correlation); by contrast, if an index is not smooth, then two close projection angles might give very different index values (fast decay of correlations with respect to distance between angles).

Luckily, the most popular class of covariance functions, known as Matérn covariance functions (see https://arxiv.org/pdf/1708.02835.pdf on page 3) has one parameter, Theta3, which controls the decay of the covariance function, which in turns controls the smoothness of the random field. The ExaGeoStat paper calls it smoothness parameter: the higher Theta3, the smoother the random field. In our context, the higher Theta3, the smoother the index function.

This suggests the following approach:

- Evaluate the index function at some random projection angle values (our initial jellies),
- Use this sample to fit a Gaussian process with Matérn covariance using ExaGeoStatR,
- Use the fitted Theta3 parameter value as our definition of smoothness of the index function: the higher Theta3, the smoother the index.

>>>>

The jellyfish optimiser (JSO) mimics the natural movements of jellyfish,
which include passive and active motions driven by ocean currents and
their swimming patterns, respectively. In the context of optimization,
these movements are abstracted to explore the search space in a way that
balances exploration (searching new areas) and exploitation (focusing on
promising areas). The algorithm aims to find the optimal solution by
adapting the jellyfish's behaviour to navigate towards the best solution
over iterations [@chou_novel_2021].

To understand what the jellyfish optimizer is doing in the context of
Projection Pursuit, we first start with a current projection (the
starting point). Then, we evaluate this projection using an index
function, which tells us how good the current projection is. We then
move the projection in a direction determined by the 'best jelly' and
random factors, influenced by how far along we are in the optimization
process (the trial $i$ and `max.tries`). Occasionally, we might explore
completely new directions like a jellyfish might with ocean currents.
Then, we compare new potential projections to our current one. If
they're better, we adopt them; if not, we stick with our current
projection. This process continues and iteratively improves the
projection, until we reach the maximum number of trials.

::: {.callout-note icon="false"}
## Algorithm: Jellyfish Optimizer Pseudo Code

@@

@@

> **if** $c_t$ is greater than or equal to $0.5$ **then**
>
> > Define trend based on the best jelly and current projections
>
> > Update each projection towards the trend using a random factor and
> > orthonormalisation
>
> **else**
>
> > **if** a random number is greater than $1 - c_t$ **then**
> >
> > > Slightly adjust each projection with a small random factor (Type A
> > > passive)
> >
> > **else**
> >
> > > For each projection, compare with a random jelly and adjust
> > > towards or away from it (Type B active)
>
> Update the orientation of each projection to maintain consistency
>
> Evaluate the new projections using the index function

> **if** any new projection is worse than the current, revert to the
> `current_projection` for that case
>
> > Determine the projection with the best index value as the new
> > best_jelly

> **if** the try is the last one, print the final best projection and
> **exit**

**return** the set of projections with the updated best_jelly as the
optimized_projection
:::

@@

```{=html}
<!--
An intuitive way to measure smoothness of a function would be to count
how many continuous derivatives exist. 
To help define smoothness in our
context, we can make use of the Sobolev spaces: functions $f$ in the
Sobolev space $W^{p,\infty}$ have all derivatives of order less than $p$
continuous. Smoothness would then be the highest $p$ such that the index
function belongs to $W^{p,\infty}$. We can relax this and consider
$W^{p,q}$ Sobolev spaces.

@@

-   $h_i$ is an neighbourhood area around $a$. We can insert different
    values of $h$ to approximate the above quantity and observe how this
    quantity changes.
-->
```
If we evaluate the index function at some random points (like the random
initialization of the jellyfish optimizer), then we can interpret these
random index values as a random field, indexed by a space parameter: the
random projection angle. This analogy suggests to use this random
training sample to fit a spatial model, a simple one being a (spatial)
Gaussian process.

```{=html}
<!--
There exist R packages to fit the hyperparameters of a Gaussian process, for example ExaGeoStatR (https://github.com/ecrc/exageostatR), which is an R wrapper for the ExaGeoStat library (https://arxiv.org/pdf/1708.02835.pdf).
-->
```

How can we define a measure of smoothness from this? The distribution of
a Gaussian process is fully determined by its mean function and
covariance function. The way the covariance function is defined is where
smoothness comes into play: if an index is very smooth, then two close
projection angles should produce close index values (strong
correlation); by contrast, if an index is not smooth, then two close
projection angles might give very different index values (fast decay of
correlations with respect to distance between angles).

Popular covariance functions are parametric positive semi-definite
functions, some of which have a parameter to capture the smoothness of
the Gaussian field. In particular, consider the Matérn class of
covariance functions, defined by

$$
K(u):=\frac{(\sqrt{2\nu}u)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}(\sqrt{2\nu}u)
$$

where $\nu>0$ is the smoothness parameter and where $\mathcal{K}_\nu$ is
the modified Bessel function. The Matérn covariance function can be
expressed analytically when $\nu$ is a half-integer, the most popular
values in the literature being $1/2$, $3/2$ and $5/2$ . The parameter
$\nu$, called smoothness parameter, controls the decay of the covariance
function. As such, it is an appropriate measure of smoothness of a
random field.

In our context, we suggest to use this parameter as a measure of the
smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as in the initial stage of the jellyfish random
search. There exist several R packages, such as GpGp or ExaGeoStatR, to
fit the hyperparameters of a GP covariance function on data. In this
project, we make use of the GpGp package.

The fitted value $\nu>0$ can be interpreted as follows: the higher
$\nu$, the smoother the index function.
